---
title: "Inspecting data and initial modeling attempts"
subtitle: "ABCD retention WG"
author: "Luis Castro-de-Araujo ^[Post-doc T32. luis.araujo@vcuhealth.org  \n]"
date: "`r format(Sys.time(), '%d %B, %Y' )`"
institute: Virginia Institute for Psychiatric and Behavioral Genetics 
lang: en
output:
  html_document:
    code_folding: hide
    highlight: zenburn
    toc: yes
    df_print: paged    
    toc_float: yes
    code_downloading: yes
---


```{r setup, results = "hide", message = F, warning = F}

c( # I use linux, this saves my time loading/installing packages
  "ProjectTemplate", "here", "stringr",
  "dplyr", "knitr", "ggplot2", "tabplot",
  "patchwork", "tableone", "tidyr", "visdat", "glmnet",
  "doParallel", "kableExtra", "dtplyr", "purrr",
  "kableExtra"
) |>
  lapply(function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x)
    }
  })

set.seed(42) # setting seed for stochastic functions
setwd("..") # here()) # needed as we are in /src, in linux here() should be used

load.project() # Loading the project
pclean()

# R options
options(
  digits = 3, # Only two decimal digits
  scipen = 999, # Remove scientific notation
  width = 100
)


# Knitr options
knitr::opts_chunk$set(
  comment = NA, # remove comment symbol
  cache.path = "../cache/", # where should I save cache?
  fig.path = "../graphs/", # where should I save figures?
  echo = T, # dont echo by default
  cache = F, # dont cache by default
  fig.width = 10, # setting the best witdth for figures
  fig.height = 7, # best height
  dpi = 300, # high dpi for publication quality,
  error = FALSE # do not interrupt in case of errors
)


cb_palette <- c(
  "#999999", "#E69F00", "#56B4E9", "#009E73",
  "#F0E442", "#0072B2", "#D55E00", "#CC79A7"
)

theme_luis <- function() {
  return_theme <- ggplot2::theme_bw(12) +
    ggplot2::theme(
      panel.border = element_rect(colour = "black"),
      legend.background = element_rect(linetype = 1, size = 0.2, colour = 1)
    )
}

# register parallel backend
cl <- makeCluster(detectCores()-3, outfile = "")
registerDoParallel(cl)
```


```{r munge}

retention  <- read.csv("~/Downloads/dataIndiv_ID_2023-03-08.csv") %>%
  filter(event == "Baseline" | event == "1 Year") %>%
  select(pguid, event, visit_status_aggr, income_household, site, race_ethnicity, education) %>%
  mutate(missed = ifelse(visit_status_aggr == "Missed (explained)" |
                         visit_status_aggr == "Missed (unexplained)", 0, 1),
         site = as.factor(site),
         race_ethnicity = factor(race_ethnicity),
         education = factor(education, ordered =T, 
                            levels = c("1st grade",
                                       "2nd grade",
                                       "3rd grade",
                                       "4th grade",
                                       "5th grade",
                                       "6th grade",
                                       "7th grade",
                                       "8th grade",
                                       "9th grade",
                                       "10th grade",
                                       "11th grade",
                                       "12th grade, no diploma",
                                       "High school graduate",
                                       "Some college, no degree",
                                       "Associate degree: Occupational, Technical, or Vocational",
                                       "Associate degree: Academic Program",
                                       "Bachelor's degree (ex. BA, AB, BS, BBS)",
                                       "Master's degree (ex. MA, MS, MEng, MEd, MBA)",
                                       "Professional School degree (ex. MD, DDS, DVN, JD)",
                                       "Doctoral degree (ex. PhD, EdD)",
                                       "GED or equivalent",
                                       "Refused to answer")),
         income_household = factor(income_household, ordered = T,
                                   levels = c("<25k",
                                              "25k-50k",
                                              "50k-75k",
                                              "75k-100k",
                                              "100k-200k",
                                              ">200k",
                                              "Don't know",
                                              "Refuse to answer")),
         # drop levels for dont know and refuse
         income_household = droplevels(income_household, c("Don't know", "Refuse to answer")),
         education = droplevels(education, c("Refused to answer")),
         income_household = as.numeric(income_household),
         education = as.numeric(education)
  ) %>% 
  drop_na(income_household, education)



# below is a version with wide format, which is the correct way, but there is an oddity in
# the data that the values for some demographic info only appear at the follow up, not at the baseline
# so I suppose the data was joined incorrectly.

# retention_wide <- dataIndiv_ID_2023.03.08 %>%
#   select(pguid, event, visit_status_aggr, income_household, site, race_ethnicity, education) %>%
#   group_by(pguid) %>%
#   filter(event == "Baseline" | event == "1 Year") %>%
#   mutate(missed = ifelse(visit_status_aggr == "Missed (explained)" |
#                 visit_status_aggr == "Missed (unexplained)", 1, 0)) %>%
#   pivot_wider(names_from = event, values_from = c(missed, income_household, site, race_ethnicity, education)) %>%
#   mutate(`missed_1 Year` = ifelse(is.na(`missed_1 Year`), 0, `missed_1 Year`))


# retention %>% 
#   group_by(event) %>% 
#   summarise(n = n(), 
#             n_missed = sum(missed_visit_1y), 
#             perc_missed = round(n_missed/n, 2)) |>
#   kable(align = "c", caption = "Number of participants and percentage of missed visits by event") |>
#   kable_styling()
# # About 5% of participants missed their 1 year visit

```

# Modeling
# Leave-one-site-out cross validation and coeff


```{r results="hide"}

# Initialize empty vectors for storing results
lambda_values <- c()
RMSE_values <- c()
var_imp_avg <- NULL

# Iterate over the sites
for (isite in unique(retention$site)) {
  # Split retention into train and test
  train <- retention %>%
    filter(site != isite)  # Leave out the current site

  test <- retention %>%
    filter(site == isite)  # Use only the current site

  # Prepare dataset in glmnet format for train and test sets
  outcome_train <- data.matrix(train$missed)
  outcome_test <- data.matrix(test$missed)

  predictors_train <- train %>%
    select(income_household,  race_ethnicity, education) %>%
    data.matrix()

  predictors_test <- test %>%
    select(income_household,  race_ethnicity, education) %>%
    model.matrix(~ ., data = .)

  message("Site: ", isite)

  # Perform cross-validated glmnet model fitting
  cv_model <- cv.glmnet(predictors_train, outcome_train,
                        nfolds = 21,  # Use LOOCV
                family = "binomial",
  parallel =T)

  # Get the best lambda
  best_lambda <- cv_model$lambda.1se

  # Fit the model using the best lambda on the test set
  fit <- glmnet(predictors_test, outcome_test,
                alpha = 1,
                type.measure = "mse",
                lambda = best_lambda,
                family = "binomial",
                intercept = FALSE)

  var_imp <- broom::tidy(fit)

  # Store the results
  RMSE <- cv_model$cvm[cv_model$lambda == cv_model$lambda.1se] %>% sqrt() %>% round(2)
  lambda <- cv_model$lambda.1se %>% round(2)

  if (is.null(var_imp_avg)) {
    var_imp_avg <- var_imp |>
      mutate(site = isite) 
  } else {
    # var_imp_avg <- full_join(var_imp_avg, var_imp, by = c("class","term"))
    var_imp_avg <- var_imp |>
      mutate(site = isite) |>
      bind_rows(var_imp_avg)

    lambda_values <- lambda_values %>% append(lambda)
    RMSE_values <- RMSE_values %>% append(RMSE)
  }
}

a <- var_imp_avg %>%
  # filter(class == 0) %>%
  group_by(term) %>%
  summarise(avg_coef = mean(estimate), 
            n = n(),
            std_error = sd(estimate) / sqrt(n)) %>%
  mutate(lower = avg_coef - 1.96 * std_error,
         upper = avg_coef + 1.96 * std_error
         ) %>%
  drop_na() # will be only keeping vars that consistently appear in all sites

```


# Factors  associated w returning to follow-up, consistently across sites {.tabset}

## Table 

```{r}

a %>% 
  select(term, avg_coef, lower, upper) %>%
  # decreasing avg_coef
  arrange(desc(avg_coef)) %>%
  #mutate(Importance = round(Importance, 2)) %>% 
  kableExtra::kbl(booktabs = T,
                  col.names = c("Factor", "mean coeff.", "95% CI - lower", "95% CI - upper"),
                  escape = TRUE) %>%
  kable_styling(latex_options = c("striped", "hold_position", "repeat_header"),
  full_width = F)

```

## Point-in-range plot - predicting retention 

```{r}

# generate a point-in-range plot similar to the table above
a %>%
  ungroup() %>%
  # arrange by avg_coef
  ggplot(aes(x = term, y = avg_coef)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  labs(x = "Factor", y = "Mean coefficient", title = "Point-in-range plot of the mean coefficient of each factor") +
  theme_bw() 

```

# GLMnet model performance {.tabset}
## The lambda space


```{r}
plot(cv_model)
```

## Plot last fit

It does not work too well in the Leave-one-site-out cv, figure out how to provide this plot in this case.


```{r}
plot(fit, xvar = "dev", label = TRUE)
```




# Basic descriptive characteristics

## Missingness {.tabset}

### Tabplot 

```{r}
library(tabplot)
tableplot(retention)

```

### Vizmiss

```{r}
library(visdat)
vis_miss(retention)

```

## Distribution of age by sex


Category names? 

```{r age-by-sex, echo = T}

dataIndiv_ID_2023.03.08 %>%
  filter(kbi_gender %in% c(1, 2, 3)) %>%
  ggplot(aes(as.numeric(interview_age), fill = as.factor(kbi_gender))) +
  geom_density(alpha = .3) +
  labs(
    title = "Age Male x Female", x = "Age",
    y = "Density"
  ) +
  theme_luis()


```


# System information

```{r}
project.info
```

```{r}
sessionInfo()
```

# References
